---
title:
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  echo = TRUE,
  collapse = TRUE,
  message = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

<!-- badges: start -->

[![R-CMD-check](https://github.com/NicChr/timeplyr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/NicChr/timeplyr/actions/workflows/R-CMD-check.yaml)
[![CRAN
status](https://www.r-pkg.org/badges/version/timeplyr)](https://CRAN.R-project.org/package=timeplyr)
[![Codecov test
coverage](https://codecov.io/gh/NicChr/timeplyr/graph/badge.svg)](https://app.codecov.io/gh/NicChr/timeplyr)
<!-- badges: end -->

# timeplyr

# **Fast Tidy Tools for Date and Datetime Manipulation**

This package provides a set of functions to make working with date and
datetime data much easier!

While most time-based packages are designed to work with clean and
pre-aggregate data, timeplyr contains a set of tidy tools to complete,
expand and summarise both raw and aggregate date/datetime data.

Significant efforts have been made to ensure that grouped calculations
are fast and efficient thanks to the excellent functionality within the
[collapse](https://sebkrantz.github.io/collapse/reference/collapse-package.html)
package.

## Installation

You can install and load `timeplyr` using the below code.

```{r gh-installation, message = FALSE, eval = FALSE}
# CRAN version
install.packages("timeplyr")

# Development version
remotes::install_github("NicChr/timeplyr")
```

```{r package_load}
library(timeplyr)
library(tidyverse)
library(fastplyr)
library(nycflights13)
library(lubridate)
```

# Basic examples

## Time arithmetic

```{r}
start <- dmy("01-01-2000")

start |> 
  time_add("year")
```

Internally timeplyr makes use of custom `timespan` objects for various
time manipulations.

```{r}
timespan("months"); timespan("1 month"); timespan(months(1))

timespan("weeks", 1:10)
```

# Adding timespans

```{r}

# Parsing 
start |> time_add("10 years"); start |> time_add("decade")

# Lubridate style timespans

start |> 
  time_add(years(1))

start |> time_add(dyears(1)); start |> time_add("31557600 seconds")

# timeplyr timespans

start |> 
  time_add(timespan("years", 1))
start |> 
  time_add(timespan("years", 0:10))
```

Time arithmetic involving days, weeks, months and years is equivalent to
using lubridate periods.

Any arithmetic involving units less than days, e.g. hours, minutes and
seconds, uses normal duration arithmetic in terms of how many seconds
have passed. This is equivalent to using lubridate durations.

```{r}
# Clocks go back 1 hour at exactly 2 am
dst <- dmy_hms("26-10-2025 00:00:00", tz = "GB") |> time_add("hour")

# Below 2 are equivalent
dst |> 
  time_add(timespan("hours", 0:2))
dst + dhours(0:2)

# There is currently no timeplyr way to achieve the below result 
# 1 hour forward in clock time is actually 2 literal hours forward
dst + hours(0:2)
```

Period-based hours, minutes and seconds are not as commonly used as
literal duration-based hours, minutes and seconds. Therefore timeplyr
tries to simplify things by removing the distinction and automatically
choosing periods for units larger than an hour and durations for units
smaller than a day without the user having to contemplate which to use.

```{r}

# The lubridate and timeplyr equivalents

time_add(dst, "second"); dst + dseconds(1)
time_add(dst, "minute"); dst + dminutes(1)
time_add(dst, "hour"); dst + dhours(1)
time_add(dst, "day"); dst + days(1)
time_add(dst, "week"); dst + weeks(1)
time_add(dst, "month"); dst + months(1)
time_add(dst, "year"); dst + years(1)
```

# Time differences

Time differences are accurate and simple

```{r}
start <- today()
end <- today() |>  time_add("decade")

time_diff(start, end, "years")

for (unit in .period_units){
    cat(unit, ":", time_diff(start, end, unit), "\n")
}
```

# Months and years

When months and years are involved, timeplyr rolls impossible dates
forward by default which is different to lubridate which rolls backwards
by default

```{r}
leap <- dmy("29-02-2020")

leap |> time_add("year"); leap |> add_with_rollback(years(1))
```

timeplyr handles month addition symmetrically by rolling backwards when
adding negative months

```{r}
leap |> time_subtract("year")
```

This can all be controlled through the `roll_month` argument which has
options 'xfirst', 'xlast', 'preday', 'postday', 'boundary', 'full' and
'NA'. The 'xfirst' and 'xlast' options are timeplyr specific and signify
crossed-first and crossed-last dates respectively. 'xlast' is set by
default package-wide.

The choice of rolling impossible dates can affect time difference
calculations.

```{r}
# timeplyr 
leap_almost_one_year_old <- dmy("28-02-2021")
cat("timeplyr: ", time_diff(leap, leap_almost_one_year_old, "years"), "\n", 
    "lubridate: ", interval(leap, leap_almost_one_year_old) / years(1),
    sep = "")
```

In the above example timeplyr lets leaplings be younger for an extra
day!

Monthly time differences in timeplyr are symmetric, regardless if
`x > y` or `x < y`

```{r}
time_diff(leap, dmy("28-02-2021") + days(0:1), "years")
time_diff(leap, dmy("01-03-2019") - days(0:1), "years")

# Not symmetric
interval(leap, dmy("28-02-2021") + days(0:1)) / years(1)
interval(leap, dmy("01-03-2019") - days(0:1)) / years(1)
```

# Fixed time intervals

timeplyr makes use of its own custom time intervals. These are intervals
of a fixed width. 
For example, R Dates can be thought of as fixed intervals of exactly 1 day width.
In fact if you call `time_interval()` on a Date vector this is exactly what
you get.

```{r}
time_interval(today())
```

Going into more detail, timeplyr uses the 'resolution' of the object to calculate
the default width. Here, 'resolution' is defined as the 'smallest timespan that 
differentiates two non-fractional instances in time'. 

For dates this is one day

```{r}
resolution(today())
```

For date-times this is one second

```{r}
resolution(now())
```

Another concept timeplyr introduces is 'granularity'. This is defined as 
'the smallest common time difference' which is a metric designed to 
estimate the level of detail or frequency in which the dates are recorded.

In the `flights` dataset, flight information is recorded every hour, 
or within hourly intervals, meaning that the level of detail (or granularity)
within this dataset is hourly.

```{r}
granularity(flights$time_hour)
```

To convert these implicit intervals into explicit intervals we can use
`time_cut_width` which places these hours into hourly intervals. It places
them by default into hourly intervals because it calls `granularity()` by default.

```{r}
time_cut_width(flights$time_hour) |> head()

# Identically
time_cut_width(flights$time_hour, "1 hour") |> head()
```

A common task is to places dates into larger time intervals, e.g. converting
hourly data into weekly data. 

```{r}
time_cut_width(flights$time_hour, "week") |> 
    as_tbl() |> 
    count(value)
```

To get full weeks, simply utilise the `from` arg with `floor_date()`

```{r}
time_cut_width(
    flights$time_hour, "week" , from = min(floor_date(flights$time_hour, "week"))
) |> 
    as_tbl() |> 
    count(value)
```

# Interval metadata

Some common metadata functions

```{r}
int <- time_cut_width(today() + days(0:13), timespan("weeks"))

interval_width(int)
interval_range(int)
interval_start(int)
interval_end(int)
interval_count(int)
```

# More detail

For the R veterans among you who would like more detail, `timespans` and 
`time_intervals` are both lightweight S3 objects with some very basic attributes, 
making them work quite fast in R. 

`timespans` are simply numeric vectors with a time unit attribute.

`time_intervals` are the unclassed version of the time object they are
representing, along with a `timespan` attribute to record the interval width, 
as well as an attribute to record the original class. 

There are many methods written for both objects to ensure they work seamlessly 
with most R functions.

```{r}
timespan("days", 1:3) |> unclass()
time_interval(today()) |>  unclass()
```



## Convert `ts`, `mts`, `xts`, `zoo`and `timeSeries` objects using `ts_as_tbl`

```{r}
eu_stock <- EuStockMarkets |>
  ts_as_tbl()
eu_stock
```

## Easily plot time series using `time_ggplot`

```{r}
eu_stock |>
  time_ggplot(time, value, group)
```

For the next examples we use flights departing from New York City in
2013.

```{r}
library(nycflights13)
library(lubridate)
flights <- flights |>
  mutate(date = as_date(time_hour))
```

```{r, echo=FALSE}
theme_set(theme_minimal())
```

## `time_by`

### Group your time variable by any time unit

```{r}
flights_monthly <- flights |>
  select(date, arr_delay) |>
  time_by(date, "month")

flights_monthly
```

We can then use this to create a monthly summary of the number of
flights and average arrival delay

```{r}
flights_monthly |>
  f_summarise(n = n(),
            mean_arr_delay = mean(arr_delay, na.rm = TRUE))
```

If the time unit is left unspecified, the `time` functions try to find
the highest time unit possible.

```{r}
flights |>
  time_by(time_hour)
```

#### Check for missing gaps in time

```{r missing_dates}
missing_dates(flights$date) # No missing dates
```

```{r}
time_num_gaps(flights$time_hour) # Missing hours
```

To check for regularity use `time_is_regular`

```{r}
hours <- sort(flights$time_hour)
time_is_regular(hours, "hours")
time_is_regular(hours, "hours", allow_gaps = TRUE, allow_dups = TRUE)

# By-group
time_num_gaps(flights$time_hour, g = flights$origin)
time_is_regular(flights$time_hour, g = flights$origin)
```

# Grouped rolling time functions

## By-group rolling mean over the last 3 calendar months

```{r}
eu_stock <- eu_stock |>
  mutate(date = date_decimal(time))

eu_stock |>
    mutate(month_mean = time_roll_mean(value, window = months(3), 
                                       time = date, 
                                       g = group)) |>
    time_ggplot(date, month_mean, group)
```

## By-group rolling (locf) NA fill

```{r}
# Prerequisite: Create Time series with missing values
x <- ts(c(NA, 3, 4, NA, 6, NA, NA, 8))
g <- cheapr::seq_id(c(3, 5)) # Two groups of size 3 + 5

roll_na_fill(x) # Simple locf fill
roll_na_fill(x, fill_limit = 1) # Fill up to 1 NA

roll_na_fill(x, g = g) # Very efficient on large data too
```

## `year_month` and `year_quarter`

timeplyr has its own lightweight 'yearmonth' and \`yearquarter' classes
inspired by the excellent 'zoo' and 'tsibble' packages.

```{r}
today <- today()
year_month(today)
```

The underlying data for a `year_month` is the number of months since 1
January 1970 (epoch).

```{r}
unclass(year_month("1970-01-01"))
unclass(year_month("1971-01-01"))
```

To create a sequence of 'year_months', one can use base arithmetic

```{r}
year_month(today) + 0:12
year_quarter(today) + 0:4
```

## `time_elapsed()`

Let's look at the time between consecutive flights for a specific flight
number

```{r}
set.seed(42)
flight_201 <- flights |>
  f_distinct(time_hour, flight) |>
  f_filter(flight %in% sample(flight, size = 1)) |>
  f_arrange(time_hour)

tail(sort(table(time_elapsed(flight_201$time_hour, "hours"))))
```

Flight 201 seems to depart mostly consistently every 24 hours

We can efficiently do the same for all flight numbers

```{r}
# We use fdistinct with sort as it's much faster and simpler to write
all_flights <- flights |>
  f_distinct(flight, time_hour, .sort = TRUE)
all_flights <- all_flights |>
  mutate(elapsed = time_elapsed(time_hour, g = flight, fill = 0))

# Flight numbers with largest relative deviation in time between flights
all_flights |>
  tidy_quantiles(elapsed, .by = flight, pivot = "wide") |>
  mutate(relative_iqr = p75 / p25) |>
  f_arrange(desc(relative_iqr))
```

`time_seq_id()` allows us to create unique IDs for regular sequences A
new ID is created every time there is a gap in the sequence

```{r}
flights |>
  f_select(time_hour) |>
  f_arrange(time_hour) |>
  mutate(time_id = time_seq_id(time_hour)) |>
  f_filter(time_id != lag(time_id)) |>
  f_count(hour(time_hour))
```

We can see that the gaps typically occur at 11pm and the sequence
resumes at 5am.

### Other convenience functions are included below

## `calendar()`

#### Easily join common date information to your data

```{r}
flights_calendar <- flights |>
    f_select(time_hour) |>
    reframe(calendar(time_hour))
```

Now that gaps in time have been filled and we have joined our date
table, it is easy to count by any time dimension we like

```{r}
flights_calendar |> 
  f_count(isoyear, isoweek)
flights_calendar |> 
  f_count(isoweek = iso_week(time))
flights_calendar |> 
  f_count(month_l)
```

## `time_seq()`

A lubridate version of `seq()` for dates and datetimes

```{r}
start <- dmy(31012020)
end <- start + years(1)
seq(start, end, by = "month") # Base R version
time_seq(start, end, "month") # lubridate version 
```

`time_seq()` doesn't mind mixing dates and datetimes

```{r}
time_seq(start, as_datetime(end), "2 weeks")
```

It is also vectorised

```{r}
# 3 sequences
time_seq(from = start, 
         to = end, 
         timespan("months", 1:3))
# Equivalent to 
c(time_seq(start, end, "month"),
  time_seq(start, end, "2 months"),
  time_seq(start, end, "3 months"))
```

## `time_seq_sizes()`

Vectorised function that calculates time sequence lengths

```{r}
seq_lengths <- time_seq_sizes(start, start + days(c(1, 10, 20)), 
                              timespan("days", c(1, 5, 10)))
seq_lengths

# Use time_seq_v2() if you know the sequence lengths
seqs <- time_seq_v2(seq_lengths, start, timespan("days", c(1, 5, 10)))
seqs
```

## `iso_week()`

Simple function to get formatted ISO weeks.

```{r}
iso_week(today())
iso_week(today(), day = TRUE)
iso_week(today(), year = FALSE)
```

## `time_cut()`

Create pretty time axes using `time_breaks()`

```{r}
times <- flights$time_hour
dates <- flights$date

date_breaks <- time_breaks(dates, n = 12)
time_breaks <- time_breaks(times, n = 12, time_floor = TRUE)

weekly_data <- flights |>
    time_by(date, "week",
            .name = "date") |>
    f_count()
weekly_data |>
  ggplot(aes(x = interval_start(date), y = n)) + 
  geom_bar(stat = "identity", fill = "#0072B2") + 
  scale_x_date(breaks = date_breaks, labels = scales::label_date_short()) 

flights |>
  ggplot(aes(x = time_hour)) + 
  geom_bar(fill = "#0072B2") + 
  scale_x_datetime(breaks = time_breaks, labels = scales::label_date_short())
```
