---
title:
output: github_document
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  echo = TRUE,
  collapse = TRUE,
  message = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

<!-- badges: start -->
[![R-CMD-check](https://github.com/NicChr/timeplyr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/NicChr/timeplyr/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/timeplyr)](https://CRAN.R-project.org/package=timeplyr)
[![Codecov test coverage](https://codecov.io/gh/NicChr/timeplyr/graph/badge.svg)](https://app.codecov.io/gh/NicChr/timeplyr)
<!-- badges: end --> 

# timeplyr

# **Fast Tidy Tools for Date and Datetime Manipulation**

This package provides a set of functions to make working with date and datetime data much easier!

While most time-based packages are designed to work 
with clean and pre-aggregate data, timeplyr contains a set of tidy tools to 
complete, expand and summarise both raw and aggregate date/datetime data.

Significant efforts have been made to ensure that grouped calculations are 
fast and efficient thanks to the excellent functionality within the 
[collapse](https://sebkrantz.github.io/collapse/reference/collapse-package.html) package.

## Installation

You can install and load `timeplyr` using the below code.

```{r gh-installation, message = FALSE, eval = FALSE}
# CRAN version
install.packages("timeplyr")

# Development version
remotes::install_github("NicChr/timeplyr")
```

```{r package_load}
library(timeplyr)
library(tidyverse)
library(fastplyr)
library(nycflights13)
library(lubridate)
```

# Basic examples

## Time arithmetic

```{r}
start <- dmy("01-01-2000")

start |> 
  time_add("year")
```

Internally timeplyr makes use of custom `timespan` objects for various 
time manipulations. 

```{r}
timespan("months"); timespan("1 month"); timespan(months(1))

timespan("weeks", 1:10)
```

# Additing timespans

```{r}

# Parsing 
start |> time_add("10 years"); start |> time_add("decade")

# Lubridate style timespans

start |> 
  time_add(years(1))

start |> time_add(dyears(1)); start |> time_add("31557600 seconds")

# timeplyr timespans

start |> 
  time_add(timespan("years", 1))
start |> 
  time_add(timespan("years", 0:10))
```

Time arithmetic involving days, weeks, months and years 
is equivalent to using lubridate periods.

Any arithmetic involving units less than days, e.g. hours, minutes and seconds, 
uses normal duration arithmetic in terms of how many seconds have passed.
This is equivalent to using lubridate durations.

```{r}
# Clocks go back 1 hour at exactly 2 am
dst <- dmy_hms("26-10-2025 00:00:00", tz = "GB") |> time_add("hour")

# Below 2 are equivalent
dst |> 
  time_add(timespan("hours", 0:2))
dst + dhours(0:2)

# There is currently no timeplyr way to achieve the below result 
# 1 hour forward in clock time is actually 2 literal hours forward
dst + hours(0:2)
```

Period-based hours, minutes and seconds are not as commonly used as
literal duration-based hours, minutes and seconds. 
Therefore timeplyr tries to simplify things by removing the distinction and
automatically choosing periods for units larger than an hour and durations for units
smaller than a day without the user having to contemplate which to use.

```{r}
# The lubridate and timeplyr equivalents for info
time_add(dst, "second"); dst + dseconds(1)
time_add(dst, "minute"); dst + dminutes(1)
time_add(dst, "hour"); dst + dhours(1)
time_add(dst, "day"); dst + days(1)
time_add(dst, "week"); dst + weeks(1)
time_add(dst, "month"); dst + months(1)
time_add(dst, "year"); dst + years(1)
```

## Time differences

Time differences are accurate and simple

```{r}
start <- ymd_hms("2008-10-27 10:00:00", tz = "GB")
end <- ymd_hms("2010-10-24 02:00:00", tz = "GB")

time_diff(start, end, "years")
interval(start, end) / years(1)
```


## Convert `ts`, `mts`, `xts`, `zoo`and `timeSeries` objects using `ts_as_tbl`

```{r}
eu_stock <- EuStockMarkets |>
  ts_as_tbl()
eu_stock
```

## Easily plot time series using `time_ggplot`

```{r}
eu_stock |>
  time_ggplot(time, value, group)
```


For the next examples we use flights departing from New York City in 2013.

```{r}
library(nycflights13)
library(lubridate)
flights <- flights |>
  mutate(date = as_date(time_hour))
```

```{r, echo=FALSE}
theme_set(theme_minimal())
```

## `time_by`

### Group your time variable by any time unit

```{r}
flights_monthly <- flights |>
  select(date, arr_delay) |>
  time_by(date, "month")

flights_monthly
```

We can then use this to create a monthly summary of the number of flights
and average arrival delay

```{r}
flights_monthly |>
  f_summarise(n = n(),
            mean_arr_delay = mean(arr_delay, na.rm = TRUE))
```

If the time unit is left unspecified, the `time` functions try to 
find the highest time unit possible.

```{r}
flights |>
  time_by(time_hour)
```

### We can also make use of timeplyr time intervals

```{r}
quarters <- time_cut_width(flights$date, "quarter")
is_time_interval(quarters)
interval_count(quarters)

# Or simply
flights |>
  time_by(date, "quarter") |>
  f_count()
```

#### Ensure full weeks by setting `from` to the start of the week

```{r}
start <- dmy("17-Jan-2013")
flights |> 
  mutate(week = time_cut_width(date, "weeks", from = floor_date(start, unit = "week"))) |> 
  f_count(week)
```

#### Check for missing gaps in time

```{r missing_dates}
missing_dates(flights$date) # No missing dates
```

```{r}
time_num_gaps(flights$time_hour) # Missing hours
```

To check for regularity use `time_is_regular`

```{r}
hours <- sort(flights$time_hour)
time_is_regular(hours, "hours")
time_is_regular(hours, "hours", allow_gaps = TRUE, allow_dups = TRUE)

# By-group
time_num_gaps(flights$time_hour, g = flights$origin)
time_is_regular(flights$time_hour, g = flights$origin)
```

# Grouped rolling time functions

## By-group rolling mean over the last 3 calendar months

```{r}
eu_stock <- eu_stock |>
  mutate(date = date_decimal(time))

eu_stock |>
    mutate(month_mean = time_roll_mean(value, window = months(3), 
                                       time = date, 
                                       g = group)) |>
    time_ggplot(date, month_mean, group)
```

## By-group rolling (locf) NA fill

```{r}
# Prerequisite: Create Time series with missing values
x <- ts(c(NA, 3, 4, NA, 6, NA, NA, 8))
g <- cheapr::seq_id(c(3, 5)) # Two groups of size 3 + 5

roll_na_fill(x) # Simple locf fill
roll_na_fill(x, fill_limit = 1) # Fill up to 1 NA

roll_na_fill(x, g = g) # Very efficient on large data too
```

## `year_month` and `year_quarter`

timeplyr has its own lightweight 'yearmonth' and `yearquarter' classes
inspired by the excellent 'zoo' and 'tsibble' packages.


```{r}
today <- today()
year_month(today)
```

The underlying data for a `year_month` is the number of months
since 1 January 1970 (epoch).

```{r}
unclass(year_month("1970-01-01"))
unclass(year_month("1971-01-01"))
```

To create a sequence of 'year_months', one can use base arithmetic

```{r}
year_month(today) + 0:12
year_quarter(today) + 0:4
```

## `time_elapsed()`

Let's look at the time between consecutive flights for a specific flight number

```{r}
set.seed(42)
flight_201 <- flights |>
  f_distinct(time_hour, flight) |>
  f_filter(flight %in% sample(flight, size = 1)) |>
  f_arrange(time_hour)

tail(sort(table(time_elapsed(flight_201$time_hour, "hours"))))
```

Flight 201 seems to depart mostly consistently every 24 hours

We can efficiently do the same for all flight numbers

```{r}
# We use fdistinct with sort as it's much faster and simpler to write
all_flights <- flights |>
  f_distinct(flight, time_hour, .sort = TRUE)
all_flights <- all_flights |>
  mutate(elapsed = time_elapsed(time_hour, g = flight, fill = 0))

# Flight numbers with largest relative deviation in time between flights
all_flights |>
  tidy_quantiles(elapsed, .by = flight, pivot = "wide") |>
  mutate(relative_iqr = p75 / p25) |>
  f_arrange(desc(relative_iqr))
```

`time_seq_id()` allows us to create unique IDs for regular sequences
A new ID is created every time there is a gap in the sequence

```{r}
flights |>
  f_select(time_hour) |>
  f_arrange(time_hour) |>
  mutate(time_id = time_seq_id(time_hour)) |>
  f_filter(time_id != lag(time_id)) |>
  f_count(hour(time_hour))
```

We can see that the gaps typically occur at 11pm and the sequence
resumes at 5am. 

### Other convenience functions are included below

## `calendar()`

#### Easily join common date information to your data

```{r}
flights_calendar <- flights |>
    f_select(time_hour) |>
    reframe(calendar(time_hour))
```

Now that gaps in time have been filled and we have 
joined our date table, it is easy to count by any time dimension we like

```{r}
flights_calendar |> 
  f_count(isoyear, isoweek)
flights_calendar |> 
  f_count(isoweek = iso_week(time))
flights_calendar |> 
  f_count(month_l)
```

## `.time_units`

See a list of available time units

```{r}
.time_units
```

## `age_years()`

Calculate ages (years) accurately

```{r}
age_years(dmy("28-02-2000"))
```

## `time_seq()`

A lubridate version of `seq()` for dates and datetimes

```{r}
start <- dmy(31012020)
end <- start + years(1)
seq(start, end, by = "month") # Base R version
time_seq(start, end, "month") # lubridate version 

```

`time_seq()` doesn't mind mixing dates and datetimes

```{r}
time_seq(start, as_datetime(end), "2 weeks")
```

## `time_seq_v()`

A vectorised version of `time_seq()`
Currently it is vectorised over from, to and by

```{r}
# 3 sequences
time_seq_v(from = start, 
           to = end, 
           timespan("months", 1:3))
# Equivalent to 
c(time_seq(start, end, "month"),
  time_seq(start, end, "2 months"),
  time_seq(start, end, "3 months"))
```

## `time_seq_sizes()`
Vectorised function that calculates time sequence lengths

```{r}
seq_lengths <- time_seq_sizes(start, start + days(c(1, 10, 20)), 
                              timespan("days", c(1, 5, 10)))
seq_lengths

# Use time_seq_v2() if you know the sequence lengths
seqs <- time_seq_v2(seq_lengths, start, timespan("days", c(1, 5, 10)))
seqs
```

Dealing with impossible dates and datetimes is very simple

```{r}
time_seq(start, end, "month", roll_month = "postday") # roll impossible months forward
time_seq(start, end, "month", roll_month = "NA") # no roll
```

## `iso_week()`

Simple function to get formatted ISO weeks.

```{r}
iso_week(today())
iso_week(today(), day = TRUE)
iso_week(today(), year = FALSE)
```

## `time_cut()`

Create pretty time axes using `time_breaks()`

```{r}
times <- flights$time_hour
dates <- flights$date

date_breaks <- time_breaks(dates, n = 12)
time_breaks <- time_breaks(times, n = 12, time_floor = TRUE)

weekly_data <- flights |>
    time_by(date, "week",
            .name = "date") |>
    f_count()
weekly_data |>
  ggplot(aes(x = interval_start(date), y = n)) + 
  geom_bar(stat = "identity", fill = "#0072B2") + 
  scale_x_date(breaks = date_breaks, labels = scales::label_date_short()) 

flights |>
  ggplot(aes(x = time_hour)) + 
  geom_bar(fill = "#0072B2") + 
  scale_x_datetime(breaks = time_breaks, labels = scales::label_date_short())
```

